Grip strength, mating success, and immune and energy costs in a population sample of US women and men: a registered report

### 2024-03-30 paper.v2.qmd
Committing before replacing with text from google doc

---
title: "Grip strength, mating success, and immune and energy costs in a population sample of US women and men: a registered report"
author: 
  - name: Caroline B. Smith
    affiliation:
     - id: wsu
       department: Department of Anthropology
       name: Washington State University
  - name: Edward H. Hagen
    affiliation:
     - ref: wsu
date: today
format:
  elsevier-html
csl: apa.csl
bibliography: regrep.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(psych)
library(xtable)
library(knitr)
library(kableExtra)
library(weights)
library(jtools)
library(reshape2)
library(gtsummary)
library(corrplot)
library(patchwork)
source('analysis.v2.R')
```

@@

Theory and evidence suggest that the mating benefits of muscle mass in human males trade off with costs of increased energy intake and decreased measures of native immunity, likely due to an evolutionary history of sexual selection. It is unknown if females experience a similar tradeoff. Using data from the 2013-2014 phase of the National Health and Nutrition Examination Survey (NHANES), a large nationally representative sample of the US (N = 4384), we will test whether grip strength, a proxy for upper body strength and physical formidability, is a positive predictor of self-reported measures of adult mating success (lifetime number of sexual partners, past year number of sexual partners, age at first intercourse, and current partnership status) in both males and females, controlling for numerous anthropometric, socioeconomic, hormone, health, and physical activity related potential confounds. We will also test if there is a positive relationship between grip strength and dietary energy intake, and a negative relationship between grip strength and native immune function. This study replicates an earlier study of US males [@lassek2009] and additionally tests these relationships in females.

@@

The mechanisms that have shaped human sexual dimorphism and whether they have primarily operated on male or female traits remain debated. Sexual dimorphism refers to sex differences in morphological and behavioral traits, excluding reproductive organs [@plavcan2001]. Unlike most animals, mammalian sexual size dimorphism tends to be male biased [@andersson1994]. In humans, sexual size dimorphism is observable as early as the first trimester in utero where male embryos are larger than female embryos [@Bukowski]. Body composition dimorphism is evident through childhood such that, adjusted for height, males have higher lean body mass and females have higher fat mass [@kirchengast2010]. At puberty, sexual dimorphism in size and body composition significantly increases such that in adults, adjusting for a 7-8% dimorphism for height, men have 12-25% higher body mass [@lassek2022], which is slightly greater than body mass dimorphism in gibbons, and slightly less than chimpanzees [@plavcan2012; @smith1997]. Men also have larger and stronger bones [@wells2007]. Other traits that are sexually dimorphic in humans include digit ratio, voice pitch, facial features, body and facial hair growth, and canine length.

While humans are only moderately dimorphic in terms of overall body mass, this is not the case for fat and muscle allocation, which are highly dimorphic (Puts, 2010) due to women’s copious storage of fat and male investment in muscle mass. Men have between 30% and 42% more fat-free mass [@lassek2022], 61% more overall muscle mass, and 78% more muscle mass in the upper arms. This concentrated muscle dimorphism in the arms and back translates to greater upper body strength in men than women. Studies of sex differences in strength show that men are stronger on average than women on all tests of muscle strength, but especially for tests of upper-body muscle strength, where female upper-body strength is 50-60% of male upper-body strength while lower-body strength is 60-70% of male values, and trunk strength is 60% of male values (Nuzzo, 2023). Muscle mass alone does not explain the sex difference in strength, as strength assessments are often greater in males than females even when pair-matched on muscle thickness (Kataoka et al., 2023), and likewise men have greater strength-to-body mass ratios than women (Nuzzo, 2023).

There is no evidence for a sex difference in the ability of the nervous system to drive the muscle (voluntary action) but instead differences in strength lie in muscle characteristics including mass, size, and fiber type. Men have both more absolute muscle mass and greater mass proportional to body size, and these muscles have higher volume and cross-sectional area size. These differences are greatest in the upper body. Finally, a greater proportion of male to female muscle is occupied by type II muscle fibers, which create greater force than type I (Nuzzo, 2023).

@@

Darwin defined sexual selection as selection that acts, not to fit individuals to their environment, but instead as to confer advantage over others in competition for mates (Darwin, 1871). Modern definitions distinguish pre- and post-copulatory competition (Andersson and Simmons, 2006). A trait is likely to have been shaped by sexual selection when it is sexually dimorphic, develops at sexual maturity, and affects mating success, either through attracting mates or winning access to mates and fertilizations through contests, or both [@andersson1994]. Two hypotheses for the evolution of human sexual size dimorphism and body composition dimorphism dominate the literature: 1) intrasexual selection through male contests and 2) intersexual selection through female mate choice. Both assert that men with more masculine traits had higher biological fitness over evolutionary history: intrasexual selection if more formidable men were able to physically outcompete other men for access to mates, and intersexual selection if females preferentially mated with men who displayed masculine traits. The male contest and the female choice hypotheses concur that sexual size and strength dimorphism are proximately (that is, developmentally) caused by sex differences in androgen hormones in the uterine environment and during differentiation at puberty. They differ, however, in the ultimate (functional) explanations for these differences, as discussed below.

@@

The intrasexual selection hypothesis emphasizes evidence that female mate choice was limited in the ancestral past (Puts, 2010), and that instead male fitness was determined by physical contests with other males for access to mates. Under these circumstances more physically formidable males (i.e., those having the ability to inflict costs on competitors) had higher reproductive success [Puts, 2016, Hill et al., 2017; @plavcan2012]. Therefore, human males have been sexually selected to be formidable resulting in the male bias in morphological traits like bone density, height, weight, muscle mass, and strength, as well as behavioral traits such as aggression (Archer, 2009). Particular emphasis is placed on sexual dimorphism in the upper-body due to the important role upper-body strength plays in both armed and unarmed fighting ability (Sell, et al., 2012) and where sex differences in strength and muscularity are greatest as discussed above.

In extant nonhuman primate species, sexual dimorphism in body weight and canine size are strongly associated with the degree of intrasexual competition, a relationship which, if used to infer the behavior of extinct australopithecines, suggests a high degree of intrasexual competition in ancestral species (Plavcan & van Schaik, 1997). Evidence of sexual size dimorphism in Homo is less certain, but in general shows reduced size dimorphism compared to early species [@plavcan2012].

@@

In contrast to the intrasexual selection hypothesis, others argue that sexual dimorphism in humans is largely the result of female choice, not for more formidable or competitive mates, but for high quality males. The immunocompetence handicap hypothesis relies on evidence that testosterone is immunosuppressive, and therefore only males with highly competent immune systems can afford to pay the costs of high levels of testosterone required to develop masculine secondary sexual traits (Folstad & Karter, 1992). Therefore, androgen-dependent traits could be costly signals of good genetic quality, in which case females could increase their long-term fitness by preferentially mating with males displaying these traits thus conferring heritable immunity to their offspring and securing investment from a healthy partner (Folstad & Karter, 1992). Similar arguments have been proposed which argue that masculine traits could be more general cues of quality and condition other than immunocompetence, or ability to bear other costs (Kokko et al., 2003).

In humans, the immunocompetence hypothesis has primarily been leveraged to explain male facial masculinity and voice pitch but it and more general cost models have also been argued to explain apparent female preference for muscular male body types (Frederick & Haselton, 2007). Cues of upper body strength indeed account for most of the variation in female-rated male body attractiveness (Sell et al., 2017).

A few problems arise for this hypothesis. First, evidence that testosterone is actually immunosuppressive is mixed (Nowak et al., 2018). Likewise, masculine trait expression is not consistently related to retrospective or prospective health (Boothroyd et al., 2013). Secondly, it is unclear to what extent women are attracted to testosterone-dependent traits, for instance facial masculinity. Although testosterone is associated with mating success, it does not seem to be through the mechanism of female choice for testosterone-derived face or body features since testosterone does not predict female rated masculinity or attractiveness (Peters et al., 2008).

Females might also choose males who could provide more resources, which in ancestral hunter-gatherer populations might have been physically stronger men. It could also have been the case that mate choice was heavily influenced by parents, in which case parental preferences would have been important (see Hill et al. 2017 for a brief review of alternative hypotheses for sexual strength dimorphism).

More generally, multiple mechanisms of sexual selection could have played a role in the evolution of human sexual dimorphism. Nevertheless, there is evidence that mating success due to formidability is mediated by male-rated dominance and not by female-rated attractiveness (Kordsmeyer et al., 2018; Hill et al., 2013).

@@

In a meta-analysis, Lidborg et al (2022) found that while voice pitch, height, and testosterone levels were associated with mating success in low fertility populations, only muscularity was associated with actual reproductive success in high fertility populations. The literature on the relationship between grip strength and mating and reproductive success is disproportionately reported for men (Gallup & Fink, 2018). Of the four studies which have tested the association between grip strength and sexual behavior, only two included both women and men (Gallup et al., 2007; Varella et al., 2014) and the other two only men (Shoup and Gallup, 2008; Sneade and Furnham, 2016). In all four, hand grip strength was positively related to mating success (including number of sex partners) for men, but there was either no relationship or a negative relationship for women. These four studies were in convenience samples.

There have been very few tests of the hypotheses that some female traits, such as breast size and waist-hip ratio (WHR), have evolved by sexual selection. In a meta-analysis, Lidborg and Boothroyd (2023) found weak evidence that more feminine digit ratios predicted higher fertility, insufficient evidence for voice pitch, and none regarding facial femininity, breast size or waist-hip ratio (WHR). At present there is mixed evidence that strength is related to reproductive success in women (Lidborg & Boothroyd, 2023). One study among the Himba showed that women with higher strength/muscularity had more living children and grandchildren (Atkinson et al., 2012), while a similar study among the Hadza showed no relationship between strength and reproductive outcomes (Smith et al., 2017). The dearth of research on female traits and reproductive outcomes leaves us unable to draw any firm conclusions about their selection pressures over time.

Previous research has shown that physical formidability in males (operationalized as fat free mass and limb muscle volume) predicts numbers of total and past-year self-reported sex partners, but that it also involves costs such as increased daily energy intake and decreased immune function (operationalized as C-reactive protein and white blood cell count) [@lassek2009], consistent with a tradeoff in mating success, immunity and energy costs. However, these relationships were not tested in women, for whom both selection hypotheses predict no association.

@@

The aim of this study is to replicate the costs and benefits of formidability reported by @@lassek2009 using similar nationally representative data, controlling for a wider range of possible confounds, and precisely prespecifying the statistical models in a registered report format. It also aims to investigate if these costs and benefits are also experienced by women.

In a large, nationally representative US sample, we predict that formidability will be significantly positively associated with mating success, with a significant interaction with sex such that strength will be a stronger predictor of male than female mating success. We also predict that there will be a negative association between male formidabillity and immune function, but little or no association for females. Finally, we predict that formidability will be positively associated with dietary energy intake for both males and females.

@@

The pilot study uses the 2011-2012 dataset, whereas the 2013-2014 dataset has been held out for confirmatory analysis pending in principle acceptance of this registered report. Grip strength, our key predictor variable, was only collected in these years. We include data from US adults between the ages of 18 and 60, the years when grip strength is the most stable (Hogrel, 2015). For each model, participants will be included if they have complete data for each predictor and outcome variable.

@@

Our main predictor variable is formidability, which @lassek2009 operationalized as fat-free mass and limb muscle volume. Since those variables were not available in our data, we operationalized formidability as combined grip strength (kg), the sum of the highest of three readings taken on each hand using a dynamometer. All regression models also included sex (male/female), age in years, and partnered status (not included in models where partnered status was the outcome variable). The interaction between sex and strength was crucial for our models since we want to test if the effect of strength on mating and immune outcomes is different for women than for men. Likewise, we also interacted age and partnered status with sex. We interacted age with sex since menopause occurs within the age range of our participants. For models of past year number of sexual partners we included an interaction between partnered status and grip strength since being currently partnered would plausibly impact mating behavior in the past year regardless of strength. Continuous-valued predictor variables were centered and standardized by 2 standard deviations which approximately matches the variation in binary variables like sex, thus making the regression coefficients more comparable (Gelman, 2008).

In industrialized populations like the U.S., widespread access to contraceptives uncouples reproductive success from sexual behavior. Measures of mating success, including number of sexual partners and age at first sexual intercourse, are used as proxies as they are assumed to have been strongly correlated with reproductive success under ancestral conditions (Pérusse, 1993). Our main outcome was mating success, which we operationalized in three ways following @lassek2009: 1) Total sexual partners was a count based on responses to the question: "In your lifetime, with how many men/women have you had any kind of sex?" 2) Total sexual partners in past year, from the question: "In the past 12 months, with how many men/women have you had any kind of sex?" Because of the way these questions are framed, (e.g. female participants were asked about their male partners and vice versa) these variables represent heterosexual partners (however, it is not necessarily the case that all of the participants included in these models identified as heterosexual). 3) Age at first sexual intercourse reported from: "How old were you when you had sex for the first time?" The sexual behavior questionnaire was self-administered on a computer in a private room at the examination center, using the Audio Computer Assisted Self Interview system, which allows participants to hear questions through headphones as well as read them on screen. Only respondents who could self-report were asked these questions.

Unlike @lassek2009, we also operationalized mating success in a fourth outcome measure as partnered status, as a proxy for longterm mating strategy since number of partners emphasizes success through short term mating strategies. Partnered status included participants who reported either being married or living with a partner whereas unpartnered included participants who reported being single, widowed, divorced, or separated.

Our second outcome measure was immune investment, which we operationalized as white blood cell count (1000 cells/ µL). In the original paper, @lassek2009 also used C-reactive protein (CRP) as an outcome variable, but CRP was not measured in the 2011-2014 data collection years and so is not included here.

Finally, to assess costs of greater strength, we investigated the relationship between strength and dietary energy intake. Participants reported all food and beverages they consumed in the 24 hour period prior to their interview. This interview was repeated for a different 24 hour period 3-10 days later to obtain dietary recalls for two separate days. These data were then used to estimate energy intake in the form of kilocalories (kcal) calculated by matching reported foods to the USDA's Food and Nutrient Database for Dietary Studies. For the outcome measure in this model we used the average number of calories per day calculated for each participant across their two recall days.

@@

We attempted to replicate the regression models in @lassek2009 as closely as possible. There were some unavoidable differences, however. First, our study aimed to investigate the role of sex; thus we include sex as a main effect as well as in interactions with key predictor variables. Second, some variables included in @lassek2009 were not available in our data. Third, we included additional control variables as noted below. Finally, we treated our pilot study as an exploratory study in which we fit numerous models not reported here. The models we report in the pilot study and which we will test exactly in the confirmatory study represent those that we consider to best test the hypotheses in @lassek2009 based on the theories we have described above and given our constraints.

@@

Our modelling strategy takes inspiration from @lassek2009. However, @lassek2009 utilized stepwise regression to automatically eliminate candidate predictors from their models which has since been found to overfit data and therefore estimated coefficients fail to replicate in future samples (Smith, 2018). Instead, we will fit generalized linear regression models (GLMs) with prespecified treatment and control variables chosen based on theoretical considerations and our exploratory analyses. Control variables included a much wider range of theoretically motivated potential confounds than used by @lassek2009, organized into themes. We specified five models for each mating success outcome measure in order to determine if the effect of grip strength on those outcomes was due to confounds with socioeconomic, health, hormone, or physical activity variables that have been associated with strength and sexual behavior.

@@

We first derived a simple model based on that reported in Table 2 in @lassek2009 for each outcome measure (total number of partners, past year number of partners, and age at first intercourse) predicted by age, sex, grip strength, and partnered status. This model also includes an anthropometric control, body mass index (BMI), calculated as kg/m^2, since body size could impact either strength or mating outcomes. We interacted age, strength, and BMI with sex to assess differences in these predictors for men and women.

@@

The socioeconomic control model included education and race as categorical variables, since there is evidence that race is related to variation in both strength (Johnson & Wilson, 2019) and sexual behavior (Fenton et al., 2005). Likewise, there is also variation in sexual behavior by education (Chandra et al., 2013), likewise there is variation in physical activity by education (He & Baker, 2005).

@@

An umbrella review found the grip strength is a useful indicator of general health status, early all-cause mortality, cardiovascular mortality, and disability (Soysal et al., 2021). Furthermore, health status is associated with higher likelihood of sexual activity, frequency of sexual activity, and reported quality of sex in men and women (Lindau & Gavrilova, 2010). Another study of men aged 45-59 found that lower sexual activity was associated with increased mortality (Smith et al., 1997). The health model, therefore, included a number of variables related to health. White blood cell count (1000 cells/ µL) and hemoglobin (g/dL) were included to control for acute infection. Depression, which is negatively related to strength (Smith et al., 2022), was measured using the Patient Health Questionnaire-9 (PHQ-9; Kroenke et al., 2001), a validated nine-item screening instrument. Each item represents a symptom of depression, and for each one participants were asked to consider how frequently they had been bothered by that symptom over the past two weeks, rated on a scale from 0 (not at all) to 3 (nearly everyday). These ratings were summed to produce a depression score ranging from 0-27. Chronic illness was included using Chronic Disease Score (0-6) a count of chronic diseases participants reported having been diagnosed with including diabetes, cancer, stroke, arthritis, heart disease and respiratory disease. A point was added for each disease a participant reported being diagnosed with, regardless of any impairment due to the disease. We then controlled for impairment resulting from chronic illness (Disease Impairment Score; 0-5) calculated from a different NHANES question, which asked participants to list up to five health conditions that specifically cause them to have difficulties with physical activities. We also controlled for physical disability and using the item "special equipment" which referred to participants' report that they needed special equipment to walk. Finally the 'perceived abnormal weight' variable was coded as true or false depending on whether a participant reported that they perceived their weight to be abnormal.

@@

Physical activity is related to sexual behavior and function in adults (Morris et al., 2022). Therefore, the physical activity control model included four dichotomous variables of vigorous and moderate work and recreational activity coded as 1 if participants reported that their work and/or recreation caused large increases in heart rate or breathing for at least 10 minutes continuously (vigorous work and/or rec), or small increases in breathing or heart rate for 10 minutes continuously (moderate work and/or rec), and zero if they reported their work and/or recreation did not.

>>>>

---
title: "Grip strength, mating success, and immune and energy costs in a population sample of US women and men: a registered report"
author: 
  - name: Caroline B. Smith
    affiliation:
     - id: wsu
       department: Department of Anthropology
       name: Washington State University
  - name: Edward H. Hagen
    affiliation:
     - ref: wsu
date: today
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
csl: apa.csl
bibliography: regrep.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(psych)
library(xtable)
library(knitr)
library(kableExtra)
library(weights)
library(jtools)
library(reshape2)
library(gtsummary)
library(ggcorrplot)
library(patchwork)
library(modelsummary)
source('analysis.v2.R')
```

@@

Theory and evidence suggest that the mating benefits of muscle mass in human males trade off with costs of increased energy intake and decreased measures of native immunity, likely due to an evolutionary history of sexual selection. It is unknown if females experience a similar tradeoff. Using data from the 2013-2014 phase of the National Health and Nutrition Examination Survey (NHANES), a large nationally representative sample of the US (N = 4384), we will test whether grip strength, a proxy for upper body strength and physical formidability, is a positive predictor of self-reported measures of adult mating success (lifetime number of sexual partners, past year number of sexual partners, age at first intercourse, and current partnership status) in both males and females, controlling for numerous anthropometric, socioeconomic, hormone, health, and physical activity related potential confounds. We will also test if there is a positive relationship between grip strength and dietary energy intake, and a negative relationship between grip strength and native immune function. This study replicates an earlier study of US males [@lassek2009] with numerous additional control variables, and it additionally tests these relationships in females.

@@

The mechanisms that have shaped human sexual dimorphism and whether they have primarily operated on male or female traits remain debated. Sexual dimorphism refers to sex differences in morphological and behavioral traits, excluding reproductive organs [@plavcan2001]. Unlike most animals, mammalian sexual size dimorphism, when it exists, tends to be male biased [@andersson1994; @Tombak_2024]. In humans, sexual size dimorphism is observable as early as the first trimester in utero where male embryos are larger than female embryos [@bukowski2007]. Body composition dimorphism is evident through childhood such that, adjusted for height, males have higher lean body mass and females have higher fat mass [@kirchengast2010]. At puberty, sexual dimorphism in size and body composition significantly increases such that in adults, adjusting for a 7-8% dimorphism for height, men have 12-25% higher body mass [@lassek2022], which is slightly greater than body mass dimorphism in gibbons, and slightly less than chimpanzees [@plavcan2012; @smith1997]. Men also have larger and stronger bones [@wells2007]. Other traits that are sexually dimorphic in humans include digit ratio, voice pitch, facial features, body and facial hair growth, and canine length.

While humans are only moderately dimorphic in terms of overall body mass, this is not the case for fat and muscle allocation, which are highly dimorphic [@puts2010] due to women’s copious storage of fat and male investment in muscle mass. Men have between 30% and 42% more fat-free mass [@lassek2022], 61% more overall muscle mass, and 78% more muscle mass in the upper arms. This concentrated muscle dimorphism in the arms and back translates to greater upper body strength in men than women. Studies of sex differences in strength show that men are stronger on average than women on all tests of muscle strength, but especially for tests of upper-body muscle strength, where female upper-body strength is 50-60% of male upper-body strength while lower-body strength is 60-70% of male values, and trunk strength is 60% of male values [@nuzzo2023]. Muscle mass alone does not explain the sex difference in strength, as strength assessments are often greater in males than females even when pair-matched on muscle thickness [@kataoka2023], and likewise men have greater strength-to-body mass ratios than women [@nuzzo2023].

There is no evidence for a sex difference in the ability of the nervous system to drive the muscle (voluntary action) but instead differences in strength lie in muscle characteristics including mass, size, and fiber type. Men have both more absolute muscle mass and greater mass proportional to body size, and these muscles have higher volume and cross-sectional area size. These differences are greatest in the upper body. Finally, a greater proportion of male to female muscle is occupied by type II muscle fibers, which create greater force than type I [@nuzzo2023].

@@

Two hypotheses for the evolution of human sexual size dimorphism and body composition dimorphism dominate the literature: 1) intrasexual selection through male contests and 2) intersexual selection through female mate choice. Both assert that men with more masculine traits had higher biological fitness over evolutionary history: intrasexual selection if more formidable men were able to physically outcompete other men for access to mates, and intersexual selection if females preferentially mated with men who displayed masculine traits. The male contest and the female choice hypotheses concur that sexual size and strength dimorphism are proximately (that is, developmentally) caused by sex differences in androgen hormones in the uterine environment and during differentiation at puberty. They differ, however, in the ultimate (functional) explanations for these differences, as discussed below.

@@

The intrasexual selection hypothesis emphasizes evidence that female mate choice was limited in the ancestral past [@puts2010], and that instead male fitness was determined by physical contests with other males for access to mates. Under these circumstances more physically formidable males (i.e., those having the ability to inflict costs on competitors) had higher reproductive success [@puts2016; @hill2017; @plavcan2012]. Therefore, human males have been sexually selected to be formidable resulting in the male bias in morphological traits like bone density, height, weight, muscle mass, and strength, as well as behavioral traits such as aggression [@archer2009]. Particular emphasis is placed on sexual dimorphism in the upper-body due to the important role upper-body strength plays in both armed and unarmed fighting ability [@sell2012] and where sex differences in strength and muscularity are greatest as discussed above.

In extant nonhuman primate species, sexual dimorphism in body weight and canine size are strongly associated with the degree of intrasexual competition, a relationship which, if used to infer the behavior of extinct australopithecines, suggests a high degree of intrasexual competition in ancestral species [@plavcan1997]. Evidence of sexual size dimorphism in *Homo* is less certain, but in general shows reduced size dimorphism compared to earlier hominin species [@plavcan2012].

@@

Although there is evidence in Western young adult populations that male mating success is mediated by male-rated dominance and not by female-rated attractiveness [@kordsmeyer2018; @hill2013], other researchers nevertheless argue that sexual dimorphism in humans is largely the result of female choice for high quality males [which was possibly heavily influenced by parents, in which case parental preferences would also have been important, @apostolou2007]. The immunocompetence handicap hypothesis, for example, relies on evidence that testosterone is immunosuppressive, and therefore only males with highly competent immune systems can afford to pay the costs of high levels of testosterone required to develop masculine secondary sexual traits [@folstad1992]. Androgen-dependent traits could thus be costly signals of good genetic quality, with females increasing their fitness by preferentially mating with males displaying these traits, thereby conferring heritable immunity to their offspring and securing investment from a healthy partner [@folstad1992]. Evidence that testosterone is immunosuppressive is mixed, however [@nowak2018], and although testosterone is associated with mating success, it does not seem to be through the mechanism of female choice for testosterone-derived face or body features since testosterone does not predict female rated masculinity or attractiveness [@peters2008].

Similar arguments have therefore been proposed that masculine traits could be more general cues of quality and condition, or an ability to bear other costs [@kokko2003; @frederick2007]. Cues of upper body strength indeed account for most of the variation in female-rated male body attractiveness [@sell2017]. Yet masculine trait expression is not consistently related to retrospective or prospective health [@boothroyd2013].

Females might be choosing males who could provide more resources, or men who could better protect them from attacks by other men or by predators, which in ancestral hunter-gatherer populations would have been physically stronger men [@apicella2014; @willems2017b]. There is substantial cross-cultural evidence of a female-biased preference for mates who can provide resources [@walter2020; @buss2019]. Importantly, studies of contemporary hunter-gatherers and game-theoretic models find that a mutually beneficial sexual division of labor within long-term pair bonds is critical to provisioning offspring who remain dependent on caregivers for up to 20 years or more [@davison2021; @davison2022; @alger2020; @kaplan2000]. See @fig-provisioning. Typically, forager women engage in lower risk, lower return foraging of plants and small game compatible with the cognitively demanding task of childrearing [@hagen2019], and men engage in higher risk, higher return big game hunting [@kelly2013]. This suggests that sex differences in body composition could be due, at least in part, to a sexual division of labor.

:::{#fig-provisioning}
![](Figures/plot_productivity.png)

Net energy production in chimpanzees and human foragers over the lifespan. Solid lines are females. Dashed line is Hadza males. Data from @davison2021 and @davison2022.
:::

More generally, multiple forms of sexual selection could have played a role in the evolution of human sexual dimorphism [see @hill2017 for a brief review of alternative hypotheses for sexual strength dimorphism].

@@

A meta-analysis found that whereas voice pitch, height, and testosterone levels were associated with mating success in low fertility populations, only muscularity was associated with actual reproductive success in high fertility populations [@lidborg2022]. In four studies of convenience samples (two of which included women), hand grip strength was positively related to mating success (including number of sex partners) for men, but there was either no relationship or a negative relationship for women [@gallup2007; @varella2014; @shoup2008; @sneade2016]

The literature on the relationship between grip strength and mating and reproductive success is disproportionately reported for men [@gallup2018]. There have been very few tests of the hypotheses that some female traits, such as breast size and waist-hip ratio (WHR), have evolved by sexual selection. A meta-analysis found weak evidence that more feminine digit ratios predicted higher fertility, insufficient evidence for voice pitch, and none regarding facial femininity, breast size or waist-hip ratio (WHR). At present there is mixed evidence that strength is related to reproductive success in women [@lidborg2022a]. One study among the Himba showed that women with higher strength/muscularity had more living children and grandchildren [@atkinson2012], while a similar study among the Hadza showed no relationship between strength and reproductive outcomes [@smith2017a]. The dearth of research on female traits and reproductive outcomes leaves us unable to draw any firm conclusions about selection pressures over time.

Previous research has shown that physical formidability in males (operationalized as fat free mass and limb muscle volume) predicts numbers of total and past-year self-reported sex partners, but that it also involves costs such as increased daily energy intake and decreased immune function, operationalized as C-reactive protein and white blood cell count [@lassek2009], consistent with a tradeoff in mating success, immunity, and energy costs. However, these relationships were not tested in women, for whom both sexual selection hypotheses predict no association.

@@

According to the sexual selection hypothesis, males evolved greater physical formidability than females because physical formidability increased male reproductive success more than females. The aim of this study is to replicate the costs and benefits of formidability reported by @lassek2009 using similar nationally representative data, controlling for a wider range of possible confounds, and precisely prespecifying the statistical models in a registered report format. It also aims to investigate if these costs and benefits are also experienced by women, and specifically if the effect of physical strength on mating success, a proxy for reproductive success, is greater for men than women, as the sexual selection hypothesis predicts.

@@

The pilot study uses the 2011-2012 dataset, whereas the 2013-2014 dataset has been held out for confirmatory analysis pending in principle acceptance of this registered report. Grip strength, our proxy for physical formidability and our key predictor variable, was only collected in these years. We include data from US adults between the ages of 18 and 60, the years when grip strength is the most stable [@hogrel2015]. For each model, participants will be included if they have complete data for each predictor and outcome variable.

Our main outcome was mating success. In industrialized populations like the U.S., widespread access to contraceptives uncouples reproductive success from sexual behavior. Measures of mating success, including number of sexual partners and age at first sexual intercourse, are used as proxies as they are assumed to have been strongly correlated with male reproductive success under ancestral conditions [@perusse1993]. We operationalized mating success in three ways following @lassek2009: 1) Total sexual partners was a count based on responses to the question: "In your lifetime, with how many men/women have you had any kind of sex?" 2) Total sexual partners in past year, from the question: "In the past 12 months, with how many men/women have you had any kind of sex?" Because of the way these questions are framed, (e.g. female participants were asked about their male partners and vice versa) these variables represent heterosexual partners (however, it is not necessarily the case that all of the participants included in these models identified as heterosexual). 3) Age at first sexual intercourse reported from: "How old were you when you had sex for the first time?" The sexual behavior questionnaire was self-administered on a computer in a private room at the examination center, using the Audio Computer Assisted Self Interview system, which allows participants to hear questions through headphones as well as read them on screen. Only respondents who could self-report were asked these questions.

Number of partners emphasizes success through short term mating strategies, but it is an ambiguous measure as it could also represent repeated rejection, or a tendency to exaggerate or understate the actual number. It also does not clearly index female reproductive success because multiple sexual partners are likely to have increased male fitness more than female fitness [@Lehtonen_2022]. We therefore also operationalized mating success in a fourth outcome measure as partnered status as a proxy for a long-term mating strategy that, as discussed earlier, benefits both sexes. Partnered status included participants who reported either being married or living with a partner whereas unpartnered included participants who reported being single, widowed, divorced, or separated [@lassek2009, included marital status as a control variable but not as an outcome variable].

Our second outcome measure was immune investment, which we operationalized as white blood cell count (1000 cells/µL). @lassek2009 also used C-reactive protein (CRP) as an outcome variable, but CRP was not measured in the 2011-2014 data collection years and so is not included here.

Finally, to assess costs of greater strength, we investigated the relationship between strength and dietary energy and protein intake. Participants reported all food and beverages they consumed in the 24 hour period prior to their interview. This interview was repeated for a different 24 hour period 3-10 days later to obtain dietary recalls for two separate days. These data were then used to estimate energy intake in the form of kilocalories (kcal) calculated by matching reported foods to the USDA's Food and Nutrient Database for Dietary Studies. For the outcome measure in this model we used the average number of calories per day calculated for each participant across their two recall days.

@@

Our main predictor variable of mating success is formidability, which @lassek2009 operationalized as fat-free mass and limb muscle volume. Since those variables were not available in our data, we operationalized formidability as combined grip strength (kg), the sum of the highest of three readings taken on each hand using a dynamometer. All regression models also included sex (male/female), age in years, and partnered status (not included in models where partnered status was the outcome variable). The interaction between sex and strength was crucial for our models since we want to test if the effect of strength on mating and immune outcomes is different for women than for men. Likewise, we also interacted sex with age and partnered status. We interacted age with sex since menopause occurs within the age range of our participants. For models of past year number of sexual partners we included an interaction between partnered status and grip strength since being currently partnered would plausibly impact mating behavior in the past year regardless of strength. Continuous-valued predictor variables were centered and standardized by 2 standard deviations which approximately matches the variation in binary variables like sex, thus making the regression coefficients more comparable [@gelman2008].

@@

We attempted to replicate the regression models in @lassek2009 as closely as possible. There were some unavoidable differences, however. First, our study aimed to investigate the role of sex; thus we include sex as a main effect as well as in interactions with key predictor variables. Second, some variables included in @lassek2009 were not available in our data. Third, we included additional control variables as noted below. Finally, we treated our pilot study as an exploratory study in which we fit numerous models not reported here. The models we report in the pilot study and which we will test exactly in the confirmatory study represent those that we consider to best test the hypotheses in @lassek2009 based on the theories we described above, the results of our exploratory analyses, and our constraints.

@@

Our modelling strategy takes inspiration from @lassek2009. However, @lassek2009 utilized stepwise regression to automatically eliminate candidate predictors from their models which has since been found to overfit data and therefore estimated coefficients fail to replicate in future samples [@smith2018a]. Instead, we will fit generalized linear regression models (GLMs) with prespecified treatment and control variables chosen based on theoretical considerations and our exploratory analyses. Control variables included a much wider range of theoretically motivated potential confounds than used by @lassek2009, organized into themes. We specified five models for each mating success outcome measure in order to determine if the effect of grip strength on those outcomes was due to confounds with socioeconomic, health, hormone, or physical activity variables that have been associated with strength and sexual behavior.

@@

We first derived a simple model based on that reported in Table 2 in @lassek2009 for each outcome measure (total number of partners, past year number of partners, and age at first intercourse) predicted by age, sex, grip strength, and partnered status. This model also includes an anthropometric control, body mass index (BMI), calculated as $kg/m^2$, since body size could impact either strength or mating outcomes. We interacted age, strength, and BMI with sex to assess differences in these predictors for men and women.

@@

The socioeconomic control model included education and race as categorical variables, since there is evidence that race is related to variation in both strength [@johnson2019] and sexual behavior [@fenton2005]. Likewise, there is also variation in sexual behavior by education [@chandra2013], and there is variation in physical activity by education [@he2005].

@@

An umbrella review found the grip strength is a useful indicator of general health status, early all-cause mortality, cardiovascular mortality, and disability [@soysal2021]. Furthermore, healthy status is associated with higher likelihood of sexual activity, frequency of sexual activity, and reported quality of sex in men and women [@lindau2010]. Another study of men aged 45-59 found that lower sexual activity was associated with increased mortality [@daveysmith1997]. The health model, therefore, included a number of variables related to health. White blood cell count (1000 cells/µL) and hemoglobin (g/dL) were included to control for acute infection. Depression, which is negatively related to strength [@smith2022], was measured using the Patient Health Questionnaire-9 [PHQ-9, @kroenke2001], a validated nine-item screening instrument. Each item represents a symptom of depression, and for each one participants were asked to consider how frequently they had been bothered by that symptom over the past two weeks, rated on a scale from 0 (not at all) to 3 (nearly everyday). These ratings were summed to produce a depression score ranging from 0-27. Chronic illness was included using Chronic Disease Score (0-6) a count of chronic diseases participants reported having been diagnosed with including diabetes, cancer, stroke, arthritis, heart disease and respiratory disease. A point was added for each disease a participant reported being diagnosed with, regardless of any impairment due to the disease. We then controlled for impairment resulting from chronic illness (Disease Impairment Score; 0-5) calculated from a different NHANES question, which asked participants to list up to five health conditions that specifically cause them to have difficulties with physical activities. We also controlled for physical disability and using the item "special equipment" which referred to participants' report that they needed special equipment to walk. Finally the 'perceived abnormal weight' variable was coded as true or false depending on whether a participant reported that they perceived their weight to be abnormal.

@@

Physical activity is related to sexual behavior and function in adults [@morris2022]. Therefore, the physical activity control model included four dichotomous variables of vigorous and moderate work and recreational activity coded as 1 if participants reported that their work and/or recreation caused large increases in heart rate or breathing for at least 10 minutes continuously (vigorous work and/or rec), or small increases in breathing or heart rate for 10 minutes continuously (moderate work and/or rec), and zero if they reported their work and/or recreation did not.
